The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-28 paper.tex
(>) Revisao Secao 3.3

\begin{figure}
{\small
\begin{align*}
  clear( fin~p )       &= p                   \\
  clear( p~;~q )       &= clear(p)            \\
  clear( p~@~loop~q) ) &= clear(p)            \\
  clear( p~and~q )     &= clear(p)~;~clear(q) \\
  clear( p~or~q )      &= clear(p)~;~clear(q) \\
  clear( mem(id) )     &= mem(id)             \\
  clear( \_ )          &= nop
\end{align*}
}%
%\rule{14cm}{0.37pt}
\caption{
The function $clear$ extracts $fin$ expressions in parallel and put their 
bodies in sequence.
\label{fig.formal.clear}
}
\end{figure}

@@

Most statements from \CEU map directly to those presented in the reduced syntax 
of Figure~\ref{lst.formal.syntax}, namely, the \code{if}, \code{`;'}, 
\code{loop}, \code{par/and}, and \code{par/or}.
For instance, the \code{if} in the concrete language behaves exactly like the 
$if$ in the described semantics.
However, there are some significant mismatches between \CEU and the formal 
semantics, which we provide a mapping in this section.

\begin{comment}
$
    \code{if}      \mapsto if   ,\2
    \code{';'}     \mapsto ~';' ,\2
    \code{loop}    \mapsto loop ,\2
    \code{par/and} \mapsto and  ,\2
    \code{par/or}  \mapsto or
$.
\end{comment}
Again, we are not considering side-effects, which are all mapped to the $mem$ 
semantic construct.

@@

The concrete \code{await} and \code{emit} primitives of \CEU support 
communication of values between them.
In the two-step translation of Figure~\ref{lst.map.emit.await}, we start with 
the concrete program in \CEU, which communicates the value $1$ between the 
\code{emit} and \code{await} in parallel (left-most code).
In the intermediate translation, we include the shared variable \code{e\_} to 
hold the value being communicated between the two trails in order to simplify 
the \code{emit}.
Finally, we convert the program into the equivalent in the formal semantics, 
translating side-effect statements into $mem$ expressions.
External events require a similar translation, i.e., each external event has a 
corresponding variable that is explicitly set by the environment before each 
reaction chain.

@@

The \code{TICK} event has an associated variable \code{TICK\_} carrying the 
wall-clock time elapsed between the two occurrences, as depicted by the 
two-step translation of Figure~\ref{lst.map.timers.2}.

TODO: continuar explicacao

@@

The biggest mismatch between \CEU and the formal semantics is regarding 
finalization blocks, which require more complex modifications in the program
for a proper mapping using the $fin$ semantic construct.
In the three-step translation of Figure~\ref{lst.map.fin}, we start with a 
concrete program (\code{CODE-1}) that uses a \code{finalize} to safely 
\code{\_release} the reference to \code{ptr} kept after the call to 
\code{\_hold}.
In the translation to the semantics, we first need to catch the \code{do-end} 
termination to run the finalization code.
For this, we translate the block into a \code{par/or} (\code{CODE-2}) with the 
original body in parallel with a $fin$ to run the finalization code.
This way, the $fin$ body executes whenever the \code{par/or} terminates, either 
normally (after the \code{await B}) or aborted from an outer composition.
%
However, the $fin$ still (incorrectly) executes even if the call to 
\code{\_hold} is not reached in the body due to an abort before awaking from 
the \code{await A}.
%
To deal with this issue, for each $fin$ we need a corresponding flag to keep 
track of code that needs to be finalized (\code{CODE-3}).
%
The flag is initially set to false, avoiding the finalization code to execute.
Only after the call to \code{\_hold} that we set the flag to true and enable 
the $fin$ body to execute.
%
The complete translation substitutes the side-effect operations with $mem$ 
expressions (\code{CODE-4}).

>>>>

\begin{figure}
{\small
\begin{align*}
  clear( fin~p )       &= p                   \\
  clear( p~;~q )       &= clear(p)            \\
  clear( p~@~loop~q) ) &= clear(p)            \\
  clear( p~and~q )     &= clear(p)~;~clear(q) \\
  clear( p~or~q )      &= clear(p)~;~clear(q) \\
  clear( \_ )          &= nop
\end{align*}
}%
%\rule{14cm}{0.37pt}
\caption{
The function $clear$ extracts $fin$ expressions in parallel and put their 
bodies in sequence.
\label{fig.formal.clear}
}
\end{figure}

@@

Most statements from \CEU (``concrete \CEU'') map directly to those presented 
in the reduced syntax of Figure~\ref{lst.formal.syntax} (``abstract \CEU'').
%, namely, the \code{if}, \code{`;'}, \code{loop}, \code{par/and}, and 
%\code{par/or}.
For instance, the \code{if} in the concrete language behaves exactly like the 
$if$ in the formal semantics.
However, there are some significant mismatches between the concrete and 
abstract \CEU, and we discuss appropriate mappings in this section.
%
Again, we are not considering side-effects, which are all mapped to the $mem$ 
semantic construct.

\begin{comment}
$
    \code{if}      \mapsto if   ,\2
    \code{';'}     \mapsto ~';' ,\2
    \code{loop}    \mapsto loop ,\2
    \code{par/and} \mapsto and  ,\2
    \code{par/or}  \mapsto or
$.
\end{comment}

@@

The concrete \code{await} and \code{emit} primitives support communication of 
values between them.
In the two-step translation of Figure~\ref{lst.map.emit.await}, we start with 
the concrete program in \CEU, which communicates the value $1$ between the 
\code{emit} and \code{await} in parallel (left-most code).
In the intermediate translation, we include the shared variable \code{e\_} to 
hold the value being communicated between the two trails in order to simplify 
the \code{emit}.
Finally, we convert the program into the equivalent in the abstract syntax, 
translating side-effect statements into $mem$ expressions.
External events require a similar translation, i.e., each external event has a 
corresponding variable that is explicitly set by the environment before each 
reaction chain.

@@

The \code{TICK} event has an associated variable \code{TICK\_} carrying the 
wall-clock time elapsed between the two occurrences, as depicted by the 
two-step translation of Figure~\ref{lst.map.timers.2}.
In the concrete program in the left, the variable \code{dt} will hold the 
residual delta time (as described in Section~\ref{sec.ceu.timers}) after 
awaking from the timer.
In the first step of the translation, we expand the \code{await 10ms} to a 
\code{loop} that decrements the elapsed number of microseconds for each 
occurrence of \code{TICK}.
When the variable \code{tot} reaches zero, we escape the \code{loop} setting 
the variable \code{dt} to contain the appropriate delta.
In the last step, we convert the program to the abstract syntax.

@@

The biggest mismatch between concrete and abstract \CEU is regarding 
finalization blocks, which require more complex modifications in the program
for a proper mapping using the $fin$ expressions.
In the three-step translation of Figure~\ref{lst.map.fin}, we start with a 
concrete program (\code{CODE-1}) that uses a \code{finalize} to safely 
\code{\_release} the reference to \code{ptr} kept after the call to 
\code{\_hold}.
In the translation, we first need to catch the \code{do-end} termination to run 
the finalization code.
For this, we translate the block into a \code{par/or} (\code{CODE-2}) with the 
original body in parallel with a $fin$ to run the finalization code.
This way, the $fin$ body executes whenever the \code{par/or} terminates, either 
normally (after the \code{await B}) or aborted from an outer composition.
%
However, the $fin$ still (incorrectly) executes even if the call to 
\code{\_hold} is not reached in the body due to an abort before awaking from 
the \code{await A}.
%
To deal with this issue, for each $fin$ we need a corresponding flag to keep 
track of code that needs to be finalized (\code{CODE-3}).
%
The flag is initially set to false, avoiding the finalization code to execute.
Only after the call to \code{\_hold} that we set the flag to true and enable 
the $fin$ body to execute.
%
The complete translation substitutes the side-effect operations with $mem$ 
expressions (\code{CODE-4}).
